Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-08-14 manuscript/manuscript.Rmd
revision on introduction.

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. Built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], the Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, I advocate the following three measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.

@@

While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent. One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023]. Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal. This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variability very well. To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner, while novices often feature a low planning depth [@vanopheusden2023] and a different set of perception on the causal structure of how certain behaviors will modify the raw materials.

>>>>

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology. Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here, I believe, provide a better approximation of past humans displaying extensive cultural variability as opposed to the assumption of omniscient *Homo economicus* for most social scientists in the post-Kahneman era, but particularly anthropologists and archaeologists.

As such, built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, I advocate the following three measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.

@@

While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent. One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023]. Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal. This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variability very well. To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner. However, this is not a solid assumption given that 1) experiments and modeling studies in congnitive science have shown that novices often feature a low planning depth [@vanopheusden2023] and 2) ethnographic studies demonstrated that even expert practioniers in tradional societies can have a different set of perception on the causal structure of how certain behaviors will modify the raw materials [@harris2021].
